---
title: "RBIEN SALVIAS Tutorial"
author: "Brian Maitner"
date: "February 17, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r message=F,warning=FALSE, results='hide'}

#If you haven't already installed BIEN, do so now.
install.packages(pkgs = "devtools",repos = "https://cloud.r-project.org")
library(devtools)

install_github(repo = "bmaitner/RBIEN/BIEN")
library(BIEN)
```

# Downloading SALVIAS data with BIEN
Downloading the SALVIAS data with the BIEN package is done using the function `BIEN_plot_datasource`
```{r, message=F,warning=FALSE, results='hide'}

salvias_data <- BIEN_plot_datasource("SALVIAS")

```

If you would like additional information than is included in the default query, there are a number of options, including:

* Additional taxonomic information

* Additonal plot metadata

* Information on native status of species

* Information on the political regions that a record falls in

See `?BIEN_plot_datasource` for more information

To download the SALVIAS dataset with all of the additional information:

```{r, message=F,warning=FALSE, results='hide'}

salvias_data_full <- BIEN_plot_datasource(datasource = "SALVIAS",cultivated = T,only.new.world = F,all.taxonomy = T,native.status = T,political.boundaries = T,all.metadata = T)


```

If you would additionally like the stem measurements associated with the SALVIAS database, use the function `BIEN_stem_datasource`.

```{r, message=F,warning=FALSE, results='hide'}

salvias_stem_data <- BIEN_stem_datasource("SALVIAS")

```

As with the `BIEN_plot_datasource` there are a number of additional data available.

Note, however, that setting *native.status* or *cultivated* to **TRUE** will greatly increase the amount of time the query takes.

```{r, message=F,warning=FALSE, results='hide'}

salvias_stem_data_full <- BIEN_stem_datasource(datasource = "SALVIAS",cultivated = T,only.new.world = F,all.taxonomy = T,native.status = T,political.boundaries = T,all.metadata = T)
```

